Toward theories of deliberated preferences

### 2020-08-17 simple.tex
Minor changes, and a remark about the failure of this representation of DP

It is well known that shallow preferences do not always match reflective ones.
Indeed, forming reflective preferences is recognised as a difficult problem in many decision problems, contrary, by definition, to shallow preferences. For example, \citet[p.\ 6]{colell_microeconomic_1995} write, when presenting the preference relation that is the basis for most of their famous book about Microeconomic theory: “Introspection quickly reveals how hard it is to evaluate alternatives that are far from the realm of common experience. It takes work and serious reflection to find out one’s own preferences.” 
It can be expected that some behavior highlighted by psychologists \citet{lichtenstein_construction_2006}, such as framing effects, would be rejected by the individuals themselves when pondering their decisions.

@@

Let me state how the approach proposed here differs from those proposed in the literature, and what problems it aims at solving.

@@

This creates a risk of mis-representing the reflective preferences of the individual: not only is it not known whether the individual would reflectively want to adhere to these axioms and, if so, how she would reflectively adapt her preferences when her shallow preferences exhibit related inconsistencies; but it is not even discussed how in principle this could be known. As a result, a theory postulating constraints on individual’s behavior on the basis of rational requirements is subject to the criticism addressed by \citet[p.\ 228]{dold_toward_2018} to libertarian paternalism, in that it “effectively denies the possibility that deviations from rational choice might be something to be acknowledged as reasonable”. 
That this is an important methodological problem is sometimes recognized by authors adopting this approach. \Citet{bleichrodt_making_2001}, for example, state that laundering require controversial assumptions (p. 1500), and recommend it only as a last resort, recommending discussing with the decision-maker whenever possible to elicit her reflective preferences (p. 1499). Note that nudging \citep{thaler_nudge_2009} creates a similar problem: we have to guess what the individual would like if he thought hard about the decision problem, was well-informed, and actively worked against the psychological features he himself does not consider adequate for good decision making. \TODO{Comment about akrasia: we may want to correct for features that the individual herself can’t correct.}

@@

The observation protocol goes as follows. Given $i$ and $\emptyset ≠ B \subseteq \allalts$ and $\hist$, $i$, after having been presented $\hist$, is asked to choose among a subset of elements from $B$ among which she is indifferent. Define $\emptyset ≠ c^\hist_i(B) \subseteq B$ as the subset that $i$ chooses as her indifference choice after exposure to $\hist$.
Define $\hist \ileadsto \phi_{B, a}$ iff $a \in c^\hist_i(B)$.
Define $\hist \ileadsto \phi_{B, ¬a}$ iff $a \notin c^\hist_i(B)$.

Given $a, b \in \allalts$, define $a \gpst b$ iff $\forall \set{a, b} \subseteq B: \phi_{B, ¬b} \in \gPhi$.
Define $a \gind b$ iff either $a = b$ or $\forall \set{a, b} \subseteq B: [\phi_{B, a} \in \gPhi \land \phi_{B, b} \in \gPhi] \lor [\exists c \in B \suchthat c \gpst a \land c \gpst b]$.
Define ${\gpeq} = {\gpst} \cup {\gind}$.

An empirical theory is said to be coherent iff $\forall \emptyset ≠ B \subseteq \allalts$: $¬(\forall a \in B: \phi_{B, ¬a} \in \gPhi) \land ¬(\exists a \in B \suchthat \phi_{B, a} \in \gPhi \land \phi_{B, ¬a} \in \gPhi)$. Coherence is required for validity, but, contrary to validity, is not an empirical property, in the sense that it can be verified without experiments, as it is independent of $\ileadsto$.

@@

\begin{theorem}[Deliberated preference]
	\label{th:dp}
	If $\gpeq$ is transitive and $\gamma$ is valid, let $r$ be a choice function such that, given any $B \in \powersetz[\mathit{F}]{\allalts}$, $r(B)$ contains exactly one arbitrary argument of each undominated equivalence classes of $B$, where $\gind$ determines the equivalence classes and $\gpst$ determines the undominated elements. Then, $\forall B \in \powersetz[\mathit{F}]{\allalts}$, $r(B)$ is a deliberated restricted choice set for $i$ given $B$.
\end{theorem}
\begin{proof}
	Observe that under these hypothesis, $\gpeq$ is a transitive, reflexive binary relation on $\allalts$, $\gpst$ is its asymmetric part and $\gind$ is its symmetric part. That $\gpst$ is asymmetric follows from coherence, which requires, considering $B = \set{a, b}$, that $\phi_{B, ¬a} \notin \gPhi$ or $\phi_{B, ¬b} \notin \gPhi$ (hence $b \ngpst a$ or $a \ngpst b$). Coherence itself follows from validity.

@@

	If $b \gind a$, and because $\set{a, b} \in B$, $\phi_{B, b} \in \gPhi$ (because $\nexists c \in B \suchthat c \gpst b$, otherwise $b \notin B'$).
\end{proof}
The phrasing of this theorem considers only finite bundles for simplicity. It more generally applies to bundles that are closed according to $\gpst$. \TODO{Check.}

@@

When $\gpeq$ is not transitive, this relation still permits to define choice functions that could be said to respect, in some reasonable sense, the deliberated preference of the individuals. However, this claim is more arguable, as it is generally (though by all means not unanimously) considered in the literature that a reflective preference should be transitive. Further explorations of this possibility are left for future articles.

@@

Define $C$ a set of criteria, functions $c: \allalts → \R$ measuring the performances of the alternatives according to some point of view. Let $t: \R^C$ be a set of thresholds, thus, consisting of one real threshold $t_c \in \R$ per criterion. Say that $a \in \allalts$ is satisficing iff $\forall c \in C: c(a) ≥ t_c$.

The theory $\gamma$ says that $i$ is a satisficer using some defined criteria $C$ and thresholds $t = (t_c)_{c \in C}$. The set $\gPhi$ is defined as $\set{\phi_{a > b} \suchthat a \text{ is satisficing and $b$ is not}}$. Given $a, b \in \allalts$, with $a$ satisficing and $b$ not, define $\ar_{a > b}$ as the text: “$a$ has performance $c_1(a)$ on criterion $c_1$, and it is reasonable to consider that having at least $t_{c_1}$ is satisfying enough from the point of view captured by $c_1$, (… same phrase for the other criteria…), therefore, it is a fully satisfactory alternative; whereas $b$ is not satisfactory on criteria (… enumeration of the criteria such that $c(b) < t_{c}$).” 
Define ${\gleadsto} = \set{\ar_{a > b}, \phi_{a > b}) \suchthat a \text{ is satisficing and $b$ is not}}$.
Define $\alldargs = \gbeats = \dbeats = {+} = \emptyset$: the theory does not consider any counter-argument as potentially damaging. 
This defines the theory $(\gleadsto, \gbeats, \alldargs, \dbeats, +)$.

>>>>

It is well known that shallow preferences do not always match reflective ones.
Indeed, forming reflective preferences is recognised as a difficult problem in many decision problems, contrary, by definition, to shallow preferences. For example, \citet[p.\ 6]{colell_microeconomic_1995} write, when presenting the preference relation that is the basis for most of their famous book about Microeconomic theory: “Introspection quickly reveals how hard it is to evaluate alternatives that are far from the realm of common experience. It takes work and serious reflection to find out one’s own preferences.” 
It can be expected that some behavior highlighted by psychologists \citep{kahneman_thinking_2012}, such as framing effects, would be rejected by the individuals themselves when pondering their decisions.

@@

Let me state how the approach proposed here to capture reflective preferences differs from those proposed in the literature, and what problems it aims at solving.

@@

This creates a risk of mis-representing the reflective preferences of the individual: not only is it not known whether the individual would reflectively want to adhere to these axioms and, if so, how she would reflectively adapt her preferences when her shallow preferences exhibit related inconsistencies; but it is not even discussed how in principle this could be known. As a result, a theory postulating constraints on individual’s behavior on the basis of rational requirements is subject to the criticism addressed by \citet[p.\ 228]{dold_toward_2018} to libertarian paternalism, in that it “effectively denies the possibility that deviations from rational choice might be something to be acknowledged as reasonable”. 
That this is an important methodological problem is sometimes recognized by authors adopting this approach. \Citet{bleichrodt_making_2001}, for example, state that laundering require controversial assumptions (p.\ 1500), and recommend it only as a last resort, recommending discussing with the decision-maker whenever possible to elicit her reflective preferences (p.\ 1499). Note that nudging \citep{thaler_nudge_2009} creates a similar problem: we have to guess what the individual would like if he thought hard about the decision problem, was well-informed, and actively worked against the psychological features he himself does not consider adequate for good decision making. \TODO{Comment about akrasia: we may want to correct for features that the individual herself can’t correct.}

@@

Let $I$ and $\allargs$ be any sets of individuals and arguments.

The observation protocol goes as follows. Given $i \in I$, $\emptyset ≠ B \subseteq \allalts$ and $\hist \in \allhist$, the individual $i$, after having been presented the arguments in the sequence $\hist$, is asked to choose among a subset of elements from $B$ among which she is indifferent. Define $\emptyset ≠ c^\hist_i(B) \subseteq B$ as the subset that $i$ chooses after exposure to $\hist$.
Define $\hist \ileadsto \phi_{B, a}$ iff $a \in c^\hist_i(B)$.
Define $\hist \ileadsto \phi_{B, ¬a}$ iff $a \notin c^\hist_i(B)$.

\begin{remark}
	This does not work, because $\gamma$ supporting $\phi_{\set{a, b, c, d}, ¬b}$, $\phi_{\set{a, b, c, d}, ¬c}$ and $\phi_{\set{a, b, c, d}, ¬d}$ may be valid even if $i$ never chooses $a$: $i$ may satisfy $¬b$ (by picking $c$) when given an argument for $¬b$, satisfy both $¬b$ and $¬c$ (by picking $d$) when given arguments for $¬b$ and $¬c$, and so on. The individual $i$ is not obliged to satisfy the three arguments for $¬b$, $¬c$ and $¬d$ simultaneously, as these arguments are required to count only when they are at the end of the sequence.
\end{remark}

This defines normative theories about $\allalts$. Any empirical theory for a normative theory about $\allalts$ is said to be an empirical theory about $\allalts$.

Given an empirical theory about $\allalts$ and $a, b \in \allalts$, define $a \gpst b$ iff $\forall \set{a, b} \subseteq B: \phi_{B, ¬b} \in \gPhi$.
Define $a \gind b$ iff either $a = b$ or $\forall \set{a, b} \subseteq B: [\phi_{B, a} \in \gPhi \land \phi_{B, b} \in \gPhi] \lor [\exists c \in B \suchthat c \gpst a \land c \gpst b]$.
Define ${\gpeq} = {\gpst} \cup {\gind}$.

An empirical theory about $\allalts$ is said to be coherent iff $\forall \emptyset ≠ B \subseteq \allalts$: $¬(\forall a \in B: \phi_{B, ¬a} \in \gPhi) \land ¬(\exists a \in B \suchthat \phi_{B, a} \in \gPhi \land \phi_{B, ¬a} \in \gPhi)$. Coherence is required for validity. Note that it is not an empirical property, in the sense that it can be verified without experiments, as it is independent of $\ileadsto$.

@@

\begin{theorem}[Deliberated preference]
	\label{th:dp}
	Let $\gamma$ be a valid empirical theory for a normatively adequate normative theory about $\allalts$, and such that $\gpeq$ is transitive. Let $r$ be a choice function such that, given any $B \in \powersetz[\mathit{F}]{\allalts}$, $r(B)$ contains exactly one arbitrary argument of each undominated equivalence classes of $B$, where $\gind$ determines the equivalence classes and $\gpst$ determines the undominated elements. Then, $\forall B \in \powersetz[\mathit{F}]{\allalts}$, $r(B)$ is a deliberated restricted choice set for $i$ given $B$.
\end{theorem}
\begin{proof}
	Observe that under these hypothesis, $\gpeq$ is a transitive, reflexive binary relation on $\allalts$, $\gpst$ is its asymmetric part and $\gind$ is its symmetric part. That $\gpst$ is asymmetric follows from coherence, which requires, considering $B = \set{a, b}$, that $\phi_{B, ¬a} \notin \gPhi$ or $\phi_{B, ¬b} \notin \gPhi$ (hence $b \ngpst a$ or $a \ngpst b$). Coherence itself follows from validity.

@@

	If $b \gind a$, and because $\set{a, b} \in B$, $\phi_{B, b} \in \gPhi$ (because $\nexists c \in B \suchthat c \gpst b$, otherwise $b \notin B'$).
\end{proof}
The phrasing of this theorem considers only finite bundles for simplicity. It more generally applies to bundles that are closed according to $\gpst$. \TODO{Phrase this correctly.}

@@

When $\gpeq$ is not transitive, it still permits to define choice functions that could be said to respect, in some reasonable sense, the deliberated preference of the individuals. However, this claim is more arguable, as it is generally (though by all means not unanimously) considered in the literature that a reflective preference should be transitive. Further explorations of this possibility are left for future articles.

The claims of an empirical theory about $\allalts$ can sometimes be defined on the basis of a preference ordering $\peq$. In such a case, the claims …

@@

Define $C$ a set of criteria, functions $c: \allalts → \R$, each function measuring the performances of the alternatives according to some point of view. Let $t \in \R^C$ be a set of thresholds, thus, consisting of one real valued threshold $t_c \in \R$ per criterion. Say that $a \in \allalts$ is satisfying iff $\forall c \in C: c(a) ≥ t_c$.

The theory $\gamma$ claims, intuitively, that individuals $i \in I$ are satisficers using some defined criteria $C$ and thresholds $t = (t_c)_{c \in C}$. The set $\gPhi$ is defined as $\set{\phi_{a > b} \suchthat a \text{ is satisficing and $b$ is not}}$. Given $a, b \in \allalts$, with $a$ satisficing and $b$ not, define $\ar_{a > b}$ as the text: “$a$ has performance $c_1(a)$ on criterion $c_1$, and it is reasonable to consider that having at least $t_{c_1}$ is satisfying enough from the point of view captured by $c_1$, (… same phrase for the other criteria…), therefore, it is a fully satisfactory alternative; whereas $b$ is not satisfactory on criteria (… enumeration of the criteria such that $c(b) < t_{c}$).” 
Define ${\gleadsto} = \set{\ar_{a > b}, \phi_{a > b}) \suchthat a \text{ is satisficing and $b$ is not}}$.
Define $\alldargs = \gbeats = \dbeats = {+} = \emptyset$: the theory does not consider any counter-argument as potentially damaging. 
This defines the theory $(\gleadsto, \gbeats, \alldargs, \dbeats, +)$.
